---
title: "Customizing `superb` plots"
bibliography: "../inst/REFERENCES.bib"
csl: "../inst/apa-6th.csl"
output: 
  rmarkdown::html_vignette
description: >
  This vignette shows how to fully customize your plot
  produced by superb.
vignette: >
  %\VignetteIndexEntry{Customizing `superb` plots}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

Creating a plot involves reading raw data and compiling these into summary statistics.
This step is handled by `superb` transparently. The second, more involving step, however
is to customize the plots so that it looks appealing to the readers.

In this vignette, we go rapidly over `superb` functionalities. Instead, we provide worked-out 
examples producing fully customized plots. We proceed with examples taken from 
scientific articles. The first example produces a rain-drop plot, the second a bar plot
whose origin is not zero.

In the following, we need the following libraries:

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
## Load relevant packages
library(superb)             # for superbPlot
library(ggplot2)            # for all the graphic directives
library(gridExtra)          # for grid.arrange
```

If they are not present on your computer, first upload them to your computer with
``install.packages("name of the package")``.


# Figure 2 of Hofer, Langmann, Burkart and Neubauer, 2022.

In their study, @h2022 examined who is the best judges of one's abilities. Examining self-ratings vs. 
other-ratings in six domain, they found out that we are not always the best judges. They present in
their Figure 2 a rain-cloud plot (@allen2019raincloud) illustrating the ratings.

In what follow, we discuss how this plot could be customized after its initial creation with `superb`.

As the six domains are within-subject ratings, the data must be composed of 6 columns (at least, 
there can be additional columns; they won't be illustrated herein). In case you do not have such data, the 
following subsection generates mock data.

## Generating mock data

We generate two sets of mock data from six sets of means and standard deviations:

```{r}
Astats <- data.frame(
    MNs = c(6.75, 6.00, 5.50, 6.50, 8.00, 8.75),
    SDs = c(2.00, 3.00, 3.50, 3.50, 1.25, 1.25)
)
dtaA <- apply(Astats, 1, 
    function(stat) {rnorm(100, mean=stat[1], sd=stat[2])} 
)
dtaA <- data.frame(dtaA)
colnames(dtaA) <- c("Verbal", "Numerical", "Spatial", "Creativity", "Intrapersonal", "Interpersonal")

Bstats <- data.frame(
    MNs = c(3.33, 3.00, 2.50, 3.00, 2.75, 3.50),
    SDs = c(0.25, 0.50, 0.66, 0.50, 0.25, 0.25)
)
dtaB <- apply(Bstats, 1, 
    function(stat) {rnorm(100, mean=stat[1], sd=stat[2])} 
)
dtaB <- data.frame(dtaB)
colnames(dtaB) <- c("Verbal", "Numerical", "Spatial", "Creativity", "Intrapersonal", "Interpersonal")
```

The datasets are `data.frame`s called `dtaA` and `dtaB`. Their columns names are the dependent variables, e.g., 
"Verbal", "Numerical", "Spatial", "Creativity", "Intrapersonal", "Interpersonal".

## Making the top-row plot

For convenience, we make lists of the desired colors and labels we want to appear on the x-axis:

```{r}
mycolors <- c("seagreen","chocolate2","mediumpurple3","deeppink","chartreuse4", "darkgoldenrod1")
mylabels <- c("Verbal", "Numerical", "Spatial", "Creativity", "Intrapersonal", "Interpersonal")
```

We are ready to make the plot with the desired adjustments:

```{r, message=FALSE, fig.width=6.7, fig.height=2.5, fig.cap="Figure 2, preliminary version"}
pltA <- superb(
    crange(Verbal, Interpersonal) ~ ., # no between-subject factors
    dtaA,                       # plot for the first data set...
    WSFactors   = "Domain(6)",  # ...a within-subject design with 6 levels
    adjustments = list(
        purpose = "difference", # we want to compare means
        decorrelation = "CM"    # and error bars are correlated-adjusted
    ),
    plotStyle="raincloud",

    # the following (optional) arguments are adjusting some of the visuals
    pointParams    = list(size = 0.75),
    jitterParams   = list(width =0.1, shape=21,size=0.05,alpha=1), # less dispersed jitter dots,
    violinParams   = list(trim=TRUE, alpha=1),                     # not transparent,
    errorbarParams = list(width = 0.1, linewidth=0.5)              # wider bars, thicker lines.
)
pltA
```

As seen, this plot is a standard, colorless, plot. It contains all that is needed; it is just
plain drab and the labels are generic ones (on the vertical axis and on the horizontal axis).



## Adding a color layer to the plot


Using `superb`, if there is only one factor, superb will consider that
it is the one on the x-axis and there is therefore no other layers in the plot. This is 
why the current plot is colorless.

It is possible, post-hoc, to indicate that we wish additional layers in the plot.

In the present, we want to add the `fill` and the `color` of dots layers.
These layers are to be "connected" to the sole factor in the present example (that is, `Domain`).
Consequently, the  x-axis labels, the fill color and the dot color are all redondant information 
identifying the condition.

To do this, simply add an aesthetic graphic directive to `pltA` with:

```{r, message=FALSE, fig.width=6.7, fig.height=2.5, fig.cap="Figure 2, version with colors"}
pltA + aes(fill = factor(Domain), colour = factor(Domain)) 
```


## Adding graphic directives for fine-tuning the plot

We can customize any `superb` plot by adding graphic directives one-by-one using the operator `+`,
or we can collect all the directives in a list, and add this list once.
As we have two plots with mostly the same directives, we use this second approach.

Typically, a plot is customized by picking a theme. The default `theme_bw()` is grayish, so
we move to `theme_classic()`. We also customize specific aspects of this theme with `theme()`
directives.

These changes are all collected within the list `commonstyle` below:

```{r}
commonstyle <- list(
    theme_classic(),                             # It has no background, no bounding box.

    # We customize this theme further:
    theme(axis.line=element_line(linewidth=0.50), # We make the axes thicker...
        axis.text = element_text(size = 10),      # their text bigger...
        axis.title = element_text(size = 12),     # their labels bigger...
        plot.title = element_text(size = 10),     # and the title bigger as well.
        panel.grid = element_blank(),             # We remove the grid lines
        legend.position = "none"                  # ... and we hide the side legend.
    ),

    # Finally, we place tick marks on the units
    scale_y_continuous( breaks=1:10 ),

    # set the labels to be displayed 
    scale_x_discrete(name="Domain", labels = mylabels),

    # and set colours to both colour and fill layers
    scale_discrete_manual(aesthetic =c("fill","colour"), values = mycolors)
)
```

We also changed the vertical scale (tick marks at designated positions) and the horizontal scale
with names on the tick marks (sadly, `superb` replaces them with consecutive numbers...) and
colors to fill the clouds (`fill`) and their borders (`colour`) as well as the rain drop colors.

Examining this plot with the `commonstyle` added, we get

```{r, message=FALSE, fig.width=6.7, fig.height=2.5, fig.cap="Figure 2, final version"}
finalpltA <- pltA + aes(fill = factor(Domain), colour = factor(Domain)) + 
    commonstyle +                           # all the above directive are added;
    coord_cartesian( ylim = c(1,10) ) +     # the y-axis bounds are given ;
    labs(title="A") +                       # the plot is labeled "A"...
    ylab("Self-worth relevance")            # and the y-axis label given.
finalpltA
```


## Making the second row of the figure

We do exactly the same for the second plot. We just change the data set to `dtaB` and in the last
graphic directives, using options tailored specifically to this second data set (smaller y-axis range, 
different label, etc.):

```{r, message=FALSE, fig.width=6.7, fig.height=2.5, fig.cap="Figure 2, bottom row"}
pltB <- superb(
    crange(Verbal, Interpersonal) ~ ., # no between-subject factors
    dtaB,                       # the second data set...
    WSFactors = "Domain(6)",    # ...a within-subject design with 6 levels
    adjustments = list(
        purpose = "difference", # we want to compare means
        decorrelation = "CM"    # and error bars are correlated-adjusted
    ),
    plotStyle="raincloud",
    # the following (optional) arguments are adjusting some of the visuals
    pointParams    = list(size = 0.75),
    jitterParams = list(width =0.1, shape=21,size=0.05,alpha=1), # less dispersed jitter dots,
    violinParams = list(trim=TRUE, alpha=1,adjust=3),            # not semi-transparent, smoother
    errorbarParams = list(width = 0.1, linewidth=0.5)                 # wider bars, thicker lines.
)
finalpltB <- pltB + aes(fill = factor(Domain), colour = factor(Domain)) + 
    commonstyle +                           # the following three lines are the differences:
    coord_cartesian( ylim = c(1,5) ) +      # the limits, 1 to 5, are different 
    labs(title="B") +                       # the plot is differently-labeled
    ylab("Judgment certainty")              # and the y-axis label differns.
finalpltB
```



## Combining and saving both plots

Finally, we assemble the two plots together

```{r, message=FALSE, fig.width=6.7, fig.height=5.0, fig.cap="Figure 2, final version"}
finalplt <- grid.arrange(finalpltA, finalpltB, ncol=1)
```

It can be saved with high-resolution if desired with

```{r, echo=TRUE, eval=FALSE}
ggsave( "Figure2.png",
    plot=finalplt,
    device = "png",
    dpi = 320,          # pixels per inche
    units = "cm",       # or "in" for dimensions in inches
    width = 17,         # as found in the article
    height = 13
)
```

That's it!


















# Figure 2 of Ma and Abrams, 2023.

In their study, @ma23 examined whether participants can suppress attentional deployment 
under unpredictable visual distractor attributes. They found for the first time that
observers can indeed suppress salient, unique colored, distractors even if the 
color was not known before hand.


```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE}
# load manually the data for the purpose of the vignette
cleandata <- data.frame(
  subject   = c(201,202,203,204,205,206,207,208,209,210,211,212,213,214,215,216,217,218,219,220,221,222,223,224),
  absentrt  = c(0.9069648,0.7501645,0.8143321,0.9850208,0.9279098,0.9620722,1.0160006,0.8921083,0.6041074,0.647717,0.6705584,0.9938026,0.8073152,1.079257,0.8648441,0.7923577,0.7683727,0.9004377,0.9590628,0.7619962,0.7245308,0.9070973,0.6244701,0.6991465),
  presentrt = c(0.8805836,0.7227798,0.7173632,0.9084251,0.8596929,0.8488763,0.9039185,0.867465,0.5874631,0.6320984,0.6598097,0.9046643,0.7659111,0.8824536,0.8235161,0.783525,0.6950923,0.8531382,0.8037397,0.674048,0.6987675,0.8272449,0.6298569,0.6853342),
  absentacc = c(0.984375,0.9375,0.953125,0.984375,0.875,0.859375,0.953125,0.953125,0.9375,0.921875,0.953125,0.875,0.96875,0.984375,0.84375,0.921875,0.921875,0.90625,0.953125,1,0.9375,0.984375,0.96875,0.9375),
  presentacc= c(0.984375,0.9921875,0.9765625,0.9921875,0.9375,0.9140625,0.9921875,0.9453125,0.96875,0.9609375,0.9765625,0.9375,0.984375,0.9765625,0.9765625,0.9140625,0.96875,0.9140625,0.9921875,0.9609375,0.9921875,0.9765625,0.9375,0.890625)
)
```

To proceed, first get to the authors' OSF https://osf.io/r52db and follow the 
instructions to obtain the dataframe `cleandata`. 

Because response times (RTs) were recorded in second, we convert them to milisecond:

```{r}
cleandata$absentrt = cleandata$absentrt*1000
cleandata$presentrt = cleandata$presentrt*1000
```

As a check, here is the first six lines of that data frame:

```{r}
head(cleandata)
```

Please select the colors desired for the bars:

```{r}
mycolors = c("black","lightgray")
```

In addition to the above libraries, we also need the `scales` library so
that we can modify the vertical axis of the plot. Indeed, bar charts by 
default start at zero, but for the present data (response times and 
mean accuracies), a scales which does not start from zero is more appropriate.
We then create a shift transformation function with a non-zero start $d$:

```{r}
library(scales)     # for a translated scale using trans_new()

shift_trans = function(d = 0) {
  scales::trans_new("shift", transform = function(x) x - d, inverse = function(y) y + d)
}
```

We're all set! We are ready to make the first plot, here RTs, as a function
of the presence or absence of the colored distractor. Because (a) we want
to compare the bars, we use difference-adjusted confidence intervals; (b) the
data were collected in a within-subject design, we use a correlation-adjusted
confidence intervals.

```{r, fig.width=3, fig.height=4, fig.cap="Figure 1, preliminary version"}
# defaults are means with 95% confidence intervals, so not specified
pltA <- superbPlot( cleandata,
    WSFactors = "target(2)",
    variables = c("absentrt", "presentrt"),
    adjustments = list(
        purpose = "difference", 
        decorrelation = "CM"),
    errorbarParams = list(colour = "gray35", width = 0.05)
)
pltA
```

As this is the default, the vertical axis starts at zero. Let's add
the `shift_trans` scale, limit the range to 720-900, and show breaks 
on every 20 units:

```{r, fig.width=3, fig.height=4, fig.cap="Figure 2, version with adequate vertical scale"}
# attached the shifted scale to it
pltA <- pltA + scale_y_continuous(
    trans = shift_trans(720),      # use translated bars
    limits = c(720,899),           # limit the plot range
    breaks = seq(720,880,20),      # define major ticks
    expand = c(0,0) )              # no expansions over the plotting area
pltA
```

We can do better: changing the default fonts, remove the legend, etc. We store
these graphic directives in a list because the same are used for the accuracy plot:

```{r, fig.width=3, fig.height=4, fig.cap="Figure 3, version  with theme and details adjusted"}
ornaments <- list(
    theme_classic(base_size = 14) + theme( legend.position = "none" ),
    aes(width = 0.5, fill = factor(target), colour = factor(target) ),
    scale_discrete_manual(aesthetic =c("fill","colour"), values = mycolors),
    scale_x_discrete(name="Color Singleton\nDistractor", labels = c("Absent","Present"))
)
pltA <- pltA + ornaments + ylab("Reaction time (ms)")
pltA
```

Finally, we put an indication regarding the significant result:

```{r, fig.width=3, fig.height=4, fig.cap="Figure 4, final version for RTs"}
pltA <- pltA + showSignificance( c(1,2), 870, -8, 
    "Singleton presence\nbenefit, p < .001",
    segmentParams = list(linewidth = 1))

# this is it! Check the result   
pltA
```

No need to go over all the details for the mean accuracy plot. We do all the
steps in a single command:

```{r, fig.width=3, fig.height=4, fig.cap="Figure 5, final version for mean accuracies"}
pltB <- superbPlot( cleandata,
    WSFactors = "target(2)",
    variables = c("absentacc", "presentacc"),
    adjustments = list(
        purpose = "difference", 
        decorrelation = "CM"),
    errorbarParams = list(colour = "gray35", width = 0.05)
) + 
scale_y_continuous(
    trans = shift_trans(0.9),           # use translated bars
    limits = c(0.9, 1.0),               # limit the plot range
    breaks = seq(0.90, 1.00, 0.01),     # define major ticks
    expand = c(0,0) ) +                 # remove empty space around plotting surface
ornaments + 
ylab("Accuracy (proportion correct)") +
showSignificance( c(1,2), 0.985, -0.005, 
    "Singleton presence\nbenefit, p = .010", 
    segmentParams = list(linewidth = 1) )

# this is it! Check the result   
pltB
```


Put the two plots side-by-side and save your work!

```{r, fig.width=6, fig.height=4, fig.cap="Figure 6, final version"}
finalplt <- grid.arrange(pltA, pltB, ncol=2)
#ggsave( "Figure2b.png",
#    plot=finalplt,
#    device = "png",
#    dpi = 320,          # pixels per inche
#    units = "cm",       # or "in" for dimensions in inches
#    width = 20,         # as found in the article
#    height = 15
#)
```

Regarding the information provided by `superb`:

```{r}
## superb::FYI: The HyunhFeldtEpsilon measure of sphericity per group are  1.000
## superb::FYI: All the groups' data are compound symmetric. Consider using CA.
```

note that with only two repeated measures, sphericity is always met (Epsilon = 1.00) 
so nothing to do with this comment. Compound symmetry is a weaker form of the
sphericity assumption. When compound symmetry is met, you can decorrelate the
data using either ``CM`` or ``CA``. You won't see much differences between the
two techniques, so you may as well ignore this comment.


Enjoy!


# References

