---
title: "The making-of the figures in the article"
bibliography: "../inst/REFERENCES.bib"
csl: "../inst/apa-6th.csl"
output: 
  html_document:
    toc: TRUE
description: >
  This vignette provides all the steps needed to make the figures
  from the article.
vignette: >
  %\VignetteIndexEntry{The making-of the figures in the article}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

Herein we provide the full code to realize Figures 1 to 6 that illustrate
the (submitted) article describing the ``superb`` framework.
These figures are based on ficticious data sets that are 
provided with the package superb (``dataFigure1`` to ``dataFigure4``).

Before we begin, we need to load a few packages, including ``superb`` of course:

```{r, message=FALSE, echo=TRUE, eval=TRUE}
## Load relevant packages
library(superb)             # for superbPlot
library(ggplot2)            # for all the graph directives
library(grid)               # for grid.arrange
library(gridExtra)          # for grid.text
```

If they are not present on your computer, first upload them to your computer with
``install.packages("name of the package")``.


# Making Figure 1

The purpose of Figure 1 is to illustrate the difference in error bars when the 
purpose of these measures of precision is to perform pair-wise comparisons. It 
is based on the data from ``dataFigure1``, whose columns are 

```{r, message=FALSE, echo=TRUE, eval=TRUE}
head(dataFigure1)
```

where ``id`` is just a participant identifier, ``grp`` indicate group membership (here
only group 1 and group 2), and finally, ``score`` is the dependent variable.

The first panel on the left is based on stand-alone confidence intervals and 
is obtained with :

```{r, message=FALSE, echo=TRUE, eval=TRUE, fig.width=3, fig.height=4, fig.cap="Figure 1a: Making left panel of Figure 1."}
plt1a = superbPlot(dataFigure1, 
            BSFactors   = "grp", 
            variables   = "score", 
            plotStyle   = "line" ) 
plt1a
```

Note that these *stand-alone* error bars could have been obtained by adding the argument
``adjustments=list(purpose = "single")`` but as it is the default value, it can be omitted.

The default ``theme`` to ``ggplot``s is not very attractive. Let's decorate the plot a bit!
To that end, I collected some additional ggplot directives in a list:

```{r, message=FALSE, echo=TRUE, eval=TRUE}
ornateBS = list(
    xlab("Group"), 
    ylab("Attitude towards class activities"),
    scale_x_discrete(labels=c("Collaborative\ngames", "Unstructured\nactivities")), #new!
    coord_cartesian( ylim = c(75,125) ),
    geom_hline(yintercept = 100, colour = "black", size = 0.5, linetype=2),
    theme_light(base_size = 14) +
    theme( plot.subtitle = element_text(size=12))
)
```

so that the first plot, with these ornaments and a title, is:

```{r, message=FALSE, echo=TRUE, eval=TRUE, fig.width=3, fig.height=4, fig.cap="Figure 1b: Decorating left panel of Figure 1."}
plt1a <- plt1a + ornateBS + labs(subtitle="(stand-alone)\n95% CI")
plt1a
```

The second plot is obtained in a simar fashion with just one additional argument requesting
difference-adjusted confidence intervals:

```{r, message=FALSE, echo=TRUE, eval=TRUE, fig.width=3, fig.height=4, fig.cap="Figure 1c: Making and decorating central panel of Figure 1."}
plt1b = superbPlot(dataFigure1, 
            BSFactors   = "grp", 
            variables   = "score", 
            adjustments = list(purpose = "difference"), #new!
            plotStyle   = "line" )
plt1b <- plt1b + ornateBS + labs(subtitle="Difference-adjusted\n95% CI") 
plt1b
```

Finally, the *raincloud* plot is obtained by changing the ``plotStyle`` argument:

```{r, message=FALSE, echo=TRUE, eval=TRUE, fig.width=3, fig.height=4, fig.cap="Figure 1c: Making and decorating right panel of Figure 1."}
    plt1c = superbPlot(dataFigure1, 
                BSFactors   = "grp", 
                variables   = "score", 
                adjustments = list(purpose = "difference"),
                plotStyle   = "raincloud",                         # new layout!
                violinParams = list(fill = "green", alpha = 0.2) ) # added some color to the violin
    plt1c <- plt1c + ornateBS + labs(subtitle="Difference-adjusted\n95% CI") 
    plt1c
```

All three plots are showns side-by-side with:

```{r, message=FALSE, echo=TRUE, eval=TRUE, fig.width=9, fig.height=4, fig.cap="Figure 1: The complete Figure 1."}
grid.arrange(plt1a, plt1b, plt1c, ncol=3)
```

and exported to a file, with, e.g.:

```{r, message=FALSE, echo=TRUE, eval=FALSE}
png(filename = "Figure1.png", width = 640, height = 320)
grid.arrange(plt1a, plt1b, plt1c, ncol=3)
dev.off()
```

In the end, which error bars are correct? The difference-adjusted ones
of course, as confirmed by a t test:

```{r, message=FALSE, echo=TRUE, eval=TRUE}
t.test(dataFigure1$score[dataFigure1$grp==1],
        dataFigure1$score[dataFigure1$grp==2], var.equal=TRUE)
```

There are no significant difference in these data at the .05 level, so that the error bars of
a 95% confidence interval should contain the other result, as is the case in the central panel.

Finally, you could try here the Tryon's adjustments (changing to ``adjustments = list(purpose = "tryon")``). 
However, you will notice no difference at all. Indeed, the variances are almost identical in the two groups
(`r sd(dataFigure1[dataFigure1$grp ==1,]$score)` vs. `r sd(dataFigure1[dataFigure1$grp ==2,]$score)`).



# Making Figure 2

Figure 2 is made in a similar fashion, and using the same decorations ``ornate`` as above with
just a difference for the name of the variable on the horizontal axis: 

```{r, message=FALSE, echo=TRUE, eval=TRUE}
ornateWS = list(
    xlab("Moment"),                                                #different!
    scale_x_discrete(labels=c("Pre\ntreatment", "Post\ntreatment")), #new!
    ylab("Statistics understanding"),
    coord_cartesian( ylim = c(75,125) ),
    geom_hline(yintercept = 100, colour = "black", size = 0.5, linetype=2),
    theme_light(base_size = 16) +
    theme( plot.subtitle = element_text(size=12))
)
    ```

The only difference is that the data are now within-subjects in a wide format, e.g., 

```{r}
head(dataFigure2)
```

This makes the left panel:

```{r, message=FALSE, echo=TRUE, eval=TRUE, fig.width=3, fig.height=4, fig.cap="Figure 2a: Making left panel of Figure 2."}
plt2a <- superbPlot(dataFigure2, 
            WSFactors    = "Moment(2)", 
            variables   = c("pre","post"), 
            adjustments = list(purpose = "single"),
            plotStyle   = "line" ) 
plt2a <- plt2a + ornateWS + labs(subtitle="Stand-alone\n95% CI")
plt2a
```

... and this makes the central panel, adding the ``CM`` decorelation technique:
 
```{r, message=TRUE, echo=TRUE, eval=TRUE, fig.width=3, fig.height=4, fig.cap="Figure 2b: Making central panel of Figure 2."}
plt2b <- superbPlot(dataFigure2, 
            WSFactors    = "Moment(2)", 
            variables   = c("pre","post"), 
            adjustments = list(purpose = "difference", decorrelation = "CA"), #new
            plotStyle   = "line" ) 
plt2b <- plt2b + ornateWS + labs(subtitle="Correlation and difference-\nadjusted 95% CI") 
plt2b
```

As seen, ``superbPlot`` issues relevant information (indicated with ``FYI``), 
here the correlation (there is not *group* factor). 

To get a sense of the general trends in the data, we can examine the data participants
per participants, joining their results with a line. As seen below, for most participants,
the trend is upward, suggesting a strongly reliable effect of the moment:

```{r, message=FALSE, echo=TRUE, eval=TRUE, fig.width=3, fig.height=4, fig.cap="Figure 2c: Making third panel of Figure 2."}
plt2c <- superbPlot(dataFigure2, 
            WSFactors    = "Moment(2)", 
            variables   = c("pre","post"), 
            adjustments = list(purpose = "difference", decorrelation = "CA"),
            plotStyle   = "pointindividualline" )   #new
plt2c <- plt2c + ornateWS + labs(subtitle="Correlation and difference-\nadjusted 95% CI")
plt2c 
```


Just for the exercice, we also compute the plot of the difference between 
the scores. To that end, we need different ornaments:

```{r}
ornateWS2 = list(
    xlab("Difference"),                                      #different!
    scale_x_discrete(labels=c("Pre minus Post\ntreatment")), #new!
    ylab("Statistics understanding"),
    coord_cartesian( ylim = c(-25,+25) ),
    geom_hline(yintercept = 0, colour = "black", size = 0.5, linetype=2),
    theme_light(base_size = 16) +
    theme( plot.subtitle = element_text(size=12))
)
```

We then compute the differences and make the plot:


```{r, message=FALSE, echo=TRUE, eval=TRUE, fig.width=3, fig.height=4, fig.cap="Figure 2d: Making right panel of Figure 2."}
dataFigure2$diff = dataFigure2$post - dataFigure2$pre
plt2d <- superbPlot(dataFigure2, 
            WSFactor    = "Moment(1)", 
            variables   = c("diff"), 
            adjustments = list(purpose = "single", decorrelation = "none"),
            plotStyle   = "raincloud",
            violinParams = list(fill = "green") )  #new
plt2d <- plt2d + ornateWS2 + labs(subtitle="95% CI \nof the difference")
plt2d
```

This last plot is not decorrelated and is not adjusted for difference. 
Decorrelation would not do anything as ``diff`` is a single column;
difference-adjustment is inadequate here as the difference will be 
compared to a fix value (namely 0, for zero improvment)

Assembling all four panels, we get: 

```{r, message=FALSE, echo=TRUE, eval=TRUE, fig.width=9, fig.height=4, fig.cap="Figure 2: The complete Figure 2."}
grid.arrange(plt2a, plt2b, plt2c, plt2d,  ncol=4)
```

... which can be exported to a file as usual: 

```{r, message=FALSE, echo=TRUE, eval=FALSE}
png(filename = "Figure2.png", width = 850, height = 320)
grid.arrange(plt2a, plt2b, plt2c, plt2d,  ncol=4)
dev.off()
```

Which error bars are depicting the significance of the result most aptly? The adjusted
ones seen in the central panel, as confirmed by a t-test on paired data:

```{r, message=FALSE, echo=TRUE, eval=TRUE}
t.test(dataFigure2$pre, dataFigure2$post, paired=TRUE)
```




# Making Figure 3

The novel element in Figure 3 is the fact that the participants have been
recruted by clusters of participants. 

We first adapt the ornaments for this situation:

```{r, message=FALSE, echo=TRUE, eval=TRUE}
ornateCRS = list(
    xlab("Group"), 
    ylab("Quality of policies"),
    scale_x_discrete(labels=c("From various\nfields", "From the\nsame field")), #new!
    coord_cartesian( ylim = c(75,125) ),
    geom_hline(yintercept = 100, colour = "black", size = 0.5, linetype=2),
    theme_light(base_size = 14) +
    theme( plot.subtitle = element_text(size=12))
)
```

Then, we get an unadjusted plot as usual:

```{r, message=FALSE, echo=TRUE, eval=TRUE, fig.width=8, fig.height=4, fig.cap="Figure 3a: The left panel of Figure 3."}
plt3a = superbPlot(dataFigure3, 
    BSFactors    = "grp", 
    variables   = "VD", 
    adjustments = list(purpose = "single", samplingDesign = "SRS"),
    plotStyle   = "line" )
plt3a <- plt3a + ornateCRS + labs(subtitle="Stand-alone\n95% CI") 
```

Here, the option ```samplingDesign = "SRS"``` is the default and can be omitted.

To indicate the presence of *cluster-randomized sampling*, the ``samplingDesign``
option is set to ``"CRS"`` and an additional information, ``clusterColumn``
is indicated to identify the column containing the cluster membership information:

```{r, message=TRUE, echo=TRUE, eval=TRUE, fig.width=8, fig.height=4, fig.cap="Figure 3b: The central panel of Figure 3."}
plt3b <- superbPlot(dataFigure3, 
    BSFactors    = "grp", 
    variables   = "VD", 
    adjustments = list(purpose = "difference", samplingDesign = "CRS"), #new
    plotStyle   = "line", 
    clusterColumn = "cluster" )                                       #new
plt3b <- plt3b + ornateCRS + labs(subtitle="Cluster and difference-\nadjusted 95% CI")
```

An inspection of the distribution does not make the cluster structure evident...

```{r, message=FALSE, echo=TRUE, eval=TRUE, fig.width=8, fig.height=4, fig.cap="Figure 3c: The right panel of Figure 3."}
plt3c <- superbPlot(dataFigure3, 
    BSFactors    = "grp", 
    variables   = "VD", 
    adjustments = list(purpose = "difference", samplingDesign = "CRS"),
    plotStyle   = "raincloud", 
    violinParams = list(fill = "green", alpha = 0.2),
    clusterColumn = "cluster" )
plt3c <- plt3c + ornateCRS + labs(subtitle="Cluster and difference-\nadjusted 95% CI")
```

Here is the complete Figure 3:

```{r, message=FALSE, echo=TRUE, eval=TRUE, fig.width=8, fig.height=4, fig.cap="Figure 3: The complete Figure 3."}
grid.arrange(plt3a, plt3b, plt3c, ncol=3)
```

This figure is saved as before with

```{r, message=FALSE, echo=TRUE, eval=FALSE}
png(filename = "Figure3.png", width = 640, height = 320)
grid.arrange(plt3a, plt3b, plt3c, ncol=3)
dev.off()
```

To make the correct t-test in the present case, we need a correction factor called
$\lambda$. An easy way is the following [see @cl16 for more]

```{r, message=FALSE, echo=TRUE, eval=TRUE}
res    <- t.test( dataFigure3$VD[dataFigure3$grp==1], 
                  dataFigure3$VD[dataFigure3$grp==2], 
                  var.equal=TRUE)
micc   <- mean(c(0.491335, 0.203857)) # mean ICCs per group, as given by superbPlot
lambda <- CousineauLaurencelleLambda(c(micc, 5, 5, 5, 5, 5, 5))
tcorr  <- res$statistic / lambda
pcorr  <- 1-pt(tcorr,4)

cat(paste("t-test corrected for cluster-randomized sampling: t(",
    2*(dim(dataFigure3)[1]-2),") = ", round(tcorr,3),
    ", p = ", round(pcorr,3),"\n", sep= ""))
```

As seen, the proper test is returning a coherent decision with the proper error bars.





# Making Figure 4

Figure 4 is an illustration of the impact of sampling among a finite population.

```{r, message=FALSE, echo=TRUE, eval=TRUE}
ornateBS = list(
    xlab(""), 
    ylab("Metabolic score"),
    scale_x_discrete(labels=c("Response to treatment")), #new!
    coord_cartesian( ylim = c(75,125) ),
    geom_hline(yintercept = 100, colour = "black", size = 0.5, linetype=2),
    theme_light(base_size = 16) +
    theme( plot.subtitle = element_text(size=12))
)
```

Lets do Figure 4.

```{r, message=FALSE, echo=TRUE, eval=TRUE, fig.width=3, fig.height=4, fig.cap="Figure 4a: The left panel of Figure 4."}
plt4a <- superbPlot(dataFigure4, 
    BSFactors = "group", 
    variables = "score", 
    adjustments=list(purpose = "single", popSize = Inf),
    plotStyle="line" ) 
plt4a <- plt4a + ornateBS + labs(subtitle="Stand-alone\n95% CI") 
```

The option ``popSize = Inf`` is the default; it indicates that the population is presumed
of infinite size. A finite size can be given, as

```{r, message=FALSE, echo=TRUE, eval=TRUE, fig.width=3, fig.height=4, fig.cap="Figure 4b: The central panel of Figure 3b."}
plt4b <- superbPlot(dataFigure4, 
    BSFactors = "group",
    variables = "score", 
    adjustments=list(purpose = "single", popSize = 50 ), # new!
    plotStyle="line" ) 
plt4b <- plt4b + ornateBS + labs(subtitle="Population size-\nadjusted 95% CI") 
```

We illustrate the plot along some distribution information with a violin plot:

```{r, message=FALSE, echo=TRUE, eval=TRUE, fig.width=3, fig.height=4, fig.cap="Figure 4c: The right panel of Figure 3b."}
plt4c <- superbPlot(dataFigure4, 
    BSFactors = "group",
    variables = "score", 
    adjustments=list(purpose = "single", popSize = 50 ), # new!
    plotStyle="pointjitterviolin",
    violinParams = list(fill = "green", alpha = 0.2)  ) 
plt4c <- plt4c + ornateBS + labs(subtitle="Population size-\nadjusted 95% CI") 
```

Which are reunited as usual:

```{r, message=FALSE, echo=TRUE, eval=TRUE, fig.width=9, fig.height=4, fig.cap="Figure 4: The complete Figure 4."}
plt4 = grid.arrange(plt4a, plt4b, plt4c, ncol=3)
```

...and saved with:

```{r, message=FALSE, echo=TRUE, eval=FALSE}
png(filename = "Figure4.png", width = 640, height = 320)
grid.arrange(plt4a, plt4b, plt4c, ncol=3)
dev.off()
```

The corrected t-test, performed by adjusting for the proportion of the population
examined [see @t12], confirms the presence of a significant difference: 

```{r, message=FALSE, echo=TRUE, eval=TRUE}
res = t.test(dataFigure4$score, mu=100)
tcorr = res$statistic /sqrt(1-nrow(dataFigure4)/50)
pcorr = 1-pt(tcorr,24)

cat(paste("t-test corrected for finite-population size: t(",
    nrow(dataFigure4)-1,") = ", round(tcorr,3),
    ", p = ", round(pcorr,3),"\n", sep= ""))
```

```{r, echo = FALSE, message = FALSE, results = 'hide'}
cat("this will be hidden until I decide or not to put this here.\n")

######################################################
## Make figure 5
######################################################

plt5a = superbPlot(dataFigure1, BSFactors = "grp", 
    statistic = "var",
    adjustments=list(purpose = "difference"),
    variables = c("score"), plotStyle="line" ) + 
  xlab("Group") + ylab("Variance in scores") + 
  labs(title="Difference-adjusted 95% CI\nof the variance") +
  coord_cartesian( ylim = c(0,250) ) +
  geom_hline(yintercept = 100, colour = "black", size = 0.5, linetype=2)
plt5b = superbPlot(dataFigure1, BSFactors = "grp",
    statistic = "fisherskew", 
    adjustments=list(purpose = "difference"),
    variables = c("score"), plotStyle="line" ) + 
  xlab("Group") + ylab("Skew of the scores") + 
  labs(title="Difference-adjusted 95% CI\nof the Fisher skew") +
  coord_cartesian( ylim = c(-2,+2) ) + 
  geom_hline(yintercept = 0, colour = "black", size = 0.5, linetype=2)
plt5 = grid.arrange(plt5a,plt5b,ncol=2)


######################################################
## Make figure 6
######################################################

grp1 = dataFigure1$score[dataFigure1$grp==1] # extract group 1 for later checks
grp2 = dataFigure1$score[dataFigure1$grp==2] # extract group 2 for later checks

# defines Cohen's d_1
d1 = function(X) {
  mean(X-100)/sd(X) 
} 
# quick checks
d1(grp1)
d1(grp2)

CI.d1 <- function(X, gamma = .95) {
    n      <- length(X)
    lambda <- d1(X) * sqrt( n ) # the noncentrality parameter
    tlow   <- qt(1/2 - gamma/2, df = n-1, ncp = lambda )
    thig   <- qt(1/2 + gamma/2, df = n-1, ncp = lambda )
    dlow   <- tlow / sqrt(n)
    dhig   <- thig / sqrt(n) 
    c(dlow, dhig)
}    
# quick checks
CI.d1(grp1)  
CI.d1(grp2)


plt6a <- superbPlot(dataFigure1, BSFactors = "grp", 
  statistic = "mean", errorbar = "CI",
  showPlot=T, plotStyle="line",
  adjustments = list(purpose = "difference"),
  variable = c("score") 
)+ ylab("Mean +- 95% CI") +
labs(title="Mean with difference adjusted\n95% confidence interval")

plt6b <- superbPlot(dataFigure1, BSFactors = "grp", 
  statistic = "d1", errorbar = "CI",
  showPlot=T, plotStyle="line",
  postprocessfct = "poolSDTransform",
  adjustments = list(purpose = "difference"),
  variable = c("score") 
)+ ylab("Cohen's d_1 +- 95% CI") +
labs(title="Cohen's d_1 with difference adjusted\n95% confidence interval") + 
annotation_custom(grid.text("Cohen's d_p = 0.49\n95% CI = [-0.06, +1.09]",
    x=.66,y=.75,gp=gpar(fontsize=12, col="black"),
    rot = 0)) 
plt6 <- grid.arrange(plt6a, plt6b, ncol=2)

```



# References

