---
title: "(advanced) Non-factorial within-subject designs in ``superb``"
bibliography: "../inst/REFERENCES.bib"
csl: "../inst/apa-6th.csl"
output: 
  rmarkdown::html_vignette
description: >
  This vignette shows how to handle non-factorial designs
  using superb.
vignette: >
  %\VignetteIndexEntry{(advanced) Non-factorial within-subject designs in ``superb``}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


```{r, echo = FALSE, warning=FALSE, message = FALSE, results = 'hide'}
cat("this will be hidden; use for general initializations.\n")
library(superb)
library(ggplot2)
options(superb.feedback = c('design','warnings') )
```

In this vignette, we show how to display a dataset with
a non-full factorial within-subject design. A non-factorial design
is when all the levels of a first factor are not fully crossed with
all the levels of a second factor. For example, if you have a design
$A(4) \times B(4)$, you would expect in total 16 levels in a full-factorial
design. If there are fewer than 16 levels, it would mean that some combinations
were not measured. This could be because these conditions are uninteresting or
impossible.

In a between group design, non full-factorial designs are not
problematic: the columns contains the level of the factors. If some 
levels are missing, it just results in a shorter data file. All the levels
present are fully identifiable as the levels are in these columns.

In within-subject design, this is a different matter: 
the columns must be matched to a condition. If some columns are missing, 
``superb`` cannot guess which levels are present and which are absent.
It is therefore necessary to specify what levels the columns represent.


## An example

Consider a case where participants are presented with strings of 1 to 4 letters
for which a decision must be made, e.g., *Is this a real word?*.
The strings can be composed of a variable number of letters between 1 and 4, but
also, with a variable number of upper-case letters, between 1 and 4. All
the combinations of string length and uppercase are presented within subjects.

This design cannot be a a full factorial design. Indeed, when 1 letter is presented,
it is not possible that the number of upper-case letter be 2, 3 or 4.

To handle that case, it is possible to specify a ``WSDesign`` argument. 
By default, this
argument is ``"fullfactorial"``. When not full factorial, a list is given with,
for each variables used as dependent variable, the levels of the factors
they correspond.

Consider this table with number of letters presented (from 1 to 4),
and number of upper-case letters (from 1 to all). The valid cases are

|       |Nletters  |Nuppercase |
|-------|----------|-----------|
|var1   |1         | 1         |
|var2   |2         | 1         |
|var3   |3         | 1         |
|var4   |4         | 1         |
|var5   |2         | 2         |
|var6   |3         | 2         |
|var7   |4         | 2         |
|var8   |3         | 3         |
|var9   |4         | 3         |
|var10  |4         | 4         |

There are 10 conditions (instead of the $4 \times 4$ if the design had been
full factorial). The table provide for each variables the levels of the 
two factors they correspond to.

## Informing ``superb`` of a non-full factorial within-subject design

In ``superb()``, ``superbPlot()`` or ``superbData()``, we indicate non full factorial 
design with:

```
WSFactors = c("Nletters","Nuppercase"),
variables = c("var1","var2","var3","var4","var5","var6",
                "var7","var8","var9","var10"),
WSDesign  = list(c(1,1), c(2,1), c(3,1), c(4,1), c(2,2), c(3,2),
                 c(4,2), c(3,3), c(4,3), c(4,4)),
```

## Returning to the example

let's try this with a random data set. It will simulate response times for
participants to response **Word** or **Non-word* to the presented strings.
As ``GRD()`` can only generate full factorial design, we will delete some columns
aftewards.

```{r}
Fulldta <- dta <- GRD( 
                WSFactors = c("Nletters(4)","Nuppercase(4)"),
                Effects = list("Nletters" = slope(25),
                               "Nuppercase" = slope(-25) ),
                Population = list(mean = 400, stddev = 50)
)
```

This dataset simulate an effect of the number of letters (increasing
response times by 25 ms for each additional letter) and of the number
of uppercase letters (decreasing response times by 25 ms for each
additional uppercase letter).

The full data can be plotted without any difficulty with ``superbPlot()``:

```{r, message=FALSE, echo=TRUE, fig.width = 4, fig.cap="**Figure 1**. Mean response times to say **Word** or **Non-word**."}
superb(
    crange(DV.1.1, DV.4.4) ~ .,
    Fulldta,
    WSFactors = c("Nletters(4)","Nuppercase(4)"),
    plotStyle="line"
)
```

To simulate a more realistic dataset, we must remove the impossible conditions:

```{r}
# destroying the six impossible columns
dta$DV.1.2 = NULL # e.g., the condition showing one letter with 2 upper-cases.
dta$DV.1.3 = NULL
dta$DV.1.4 = NULL
dta$DV.2.3 = NULL
dta$DV.2.4 = NULL
dta$DV.3.4 = NULL
```

after which the plot can be performed:

```{r, message=FALSE, echo=TRUE, fig.width = 4, fig.cap="**Figure 2**. Mean response times to say **Word** or **Non-word**."}
superb(
    crange(DV.1.1, DV.4.4) ~ .,
    dta,
    WSFactors = c("Nletters(4)","Nuppercase(4)"),
    WSDesign  = list(c(1,1), c(2,1), c(3,1),  c(4,1),  c(2,2),  c(3,2),
                     c(4,2), c(3,3), c(4,3),  c(4,4)),
    plotStyle="line"
)
```

Really, the only new manipulation is the use of ``WSDesign`` where for each
within-subject variable in the dataset, the levels of the corresponding factors
(in the order they were given in ``WSFactors``) are listed.

(note that the WSDesign levels superseeds the total number of levels, *4*, given
in the ``WSFactors``. Hence, ``Nletters`` could be given as ``Nletters(99)`` or 
as ``Nletters(1)``, that has no longer any impact on the plot.)

## In summary

Full-factorial design are generally prefered as powerfull analyses
can manipulate such data (for example, the ANOVA). Further,
causality in experimental designs are more aptly assessed with 
the full-factorial design. However, it is not always possible 
or usefull to run a full-factorial design.

