---
title: "Devising custom plot layouts"
output: rmarkdown::html_vignette
description: >
  superbPlot can generate either line plots, barcharts, or point plots. 
  However, additional layout can be programmed. In this vignette, we 
  show how to add custom layouts in superbPlot. As an example,
  we show how to implement additional layouts.
vignette: >
  %\VignetteIndexEntry{Devising custom plot layouts}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


```{r, echo = FALSE, message = FALSE, results = 'hide'}
cat("this will be hidden; use for general initializations.")
library(superb)
library(ggplot2)
```

``superbPlot`` comes with six built-in layouts for plotting your data.
However, it is possible to add additional, custom-made layouts. In this
vignette, we present rapidly the existing layouts, then show how
to supplement ``superb`` with your own layouts.


## The built-in plot layouts


When calling ``superbPlot``, you use the ``plotStyle = "layout"`` option
to indicate which layout you wish to use. Internally, ``superbPlot``
is calling a function whose name is ``superbPlot."layout"``. For example,
with ``plotStyle = "line"``, the plot is actually performed by the 
function ``superbPlot.line``.


The six layout available in  ``superbPlot`` package are :

* ``superbPlot.line`` : shows the results as points and lines, 

* ``superbPlot.point``: shows the results as points only, 

* ``superbPlot.bar`` : shows the results using bars, 

* ``superbPlot.pointjitter``: shows the results with points, and the raw data with jittered points, 

* ``superbPlot.pointjitterviolin``: also shows violin plot behind the jitter points, and 

* ``superbPlot.pointindividualline``: show the results with fat points, and individual results with thin lines, 

To determine if a certain fonction is ``superbPlot``-compatible, 
use the following function:

```{r}
superb:::is.superbPlot.function("superbPlot.line")
```

where you put between quote the name of a function. When devising your
own, custom-made function, it is a gook thing to check that it is
superbPlot-compatible.


## Devising a custom-made plot layout

The purpose of ``superbPlot`` is to 

* compile the summary information
(location of the summary statistic, upper width and lower width of 
the interval) and that, for each level of the factors; 

* applies all the adjustments needed in producing the summary;

* ``superbPlot`` also transform to discret factors (using 
``as.factor``) all the factors. If some of the factors are
within-subject, the information is to be found in distinct column (
remember that ``superbPlot`` assumes a wide-format data frame); for
between-group factors, a column indicate the current group; 

* and finally, calls the plot function accepting pre-defined 
arguments

In devising your own plot function, it is important that the
function accept very specific arguments with very precise names.

Here is the header for a function corresponding to a
plot style called, say, *foo* (``plotStyle = "foo"``):

```{r}
superbPlot.foo <- function(
    summarydata,
    xvar,       
    groupingfac,
    addfactors, 
    Debug, 
    rawdata        
    # any optional argument you wish
) {
    plot <- ggplot() ## ggplot instructions...        
    return(plot)
}
```

(in what follow, I assume that one factor is placed on the
horizontal axis, another one is used to group the point,
and up to two additional factors will results in columns
and rows of panels. Of course, in devising your own template,
you may use different placement.) 


The arguments are:
* ``summarydata``: this data frame will contain the column ``center``
indicating the statistic's value, ``lowerwidth`` and ``upperwidth``
indicating how many units below and above ``center`` the error bar
extends. The data frame will also have columns for all the factors, 
and there will be as many lines as there are combinaisons of factors.

* ``xvar`` is the factor to put on the horizontal axis;

* ``groupingfac`` is the factor used to create groups of points;

* ``addfactors`` are up to two additional factors to create the 
rows and columns of panels. ``addfactors`` is formatted for facetting
(e.g., for factors "A" and "B", ``addfactors`` would be "A~B");

* ``Debug`` (a boolean) can be used when testing your code; it is
set to TRUE with the ``superbPlot`` argument ``Debug = TRUE``;

* ``rawdata`` : this data.frame contains the raw data with factors
being transformed ``as.factor`` and the dependent column being renamed
``DV``. When the data are in wide format, ``rawdata`` is reshaped to
long format.

* ``{optional arguments}`` can be used. They must be named here; when 
calling ``superbPlot``, any argument whose name match your optional
argument will be transmitted to your custom-made function.


## The simplest example


What follow is a simple example that will design a template 
that we will call ``simple``. This layout will display the
descriptive statistics and error bars. Everything will be
black and white (no color instruction) and superimposed
(no grouping instruction).

The result will be:

```{r, message=FALSE, echo=FALSE, fig.cap="Figure 1: Mean score with 95% confidence interval using the ``simple`` plot layout."}
superbPlot.simple <- function(
    summarydata, xvar, groupingfac, addfactors, Debug, rawdata 
) {
    plot <- ggplot(
        data = summarydata, 
        mapping = aes_string( x = xvar, y = "center")
    ) +
    geom_point( ) +
    geom_errorbar( mapping = aes_string(ymin = "center + lowerwidth", ymax = "center + upperwidth")  )+ 
    facet_grid( addfactors )
        
    return(plot)
}
superbPlot(TMB1964r,
     WSFactor = "T(7)",      
     BSFactor = "Condition",
     variables = c("T1","T2","T3","T4","T5","T6","T7"),
     plotStyle = "simple", Quiet = TRUE, Debug=T
)
```

To make this plot, we design a function ``superbPlot.simple`` as:

```{r}
superbPlot.simple <- function(
    summarydata, xvar, groupingfac, addfactors, Debug, rawdata 
) {
    plot <- ggplot(
        data = summarydata, 
        mapping = aes_string( x = xvar, y = "center")
    ) +
    geom_point( ) +
    geom_errorbar( mapping = aes_string(ymin = "center + lowerwidth", ymax = "center + upperwidth")  )+ 
    facet_grid( addfactors )
        
    return(plot)
}
```

The first instruction, ``ggplot`` defines the source data to be 
``summarydata`` with horizontal axis being in the string ``xvar``
(this is the reason that ``mapping`` must be given with ``aes_string``).
The position of the descriptive statistics will be in ``"center"``.

The second instruction put points for each ``"center"``, and 
the third instruction places error bars. In that case, the ``ymin`` and 
``ymax`` information are contained in ``center+lowerwidth`` and
``center+upperwidth`` where ``lowerwidth`` and ``upperwidth`` are
in the summarydata dataframe.

The last instructions generates distinct panels for each level of the
remaining factors.
 
You can check that this function is ``superbPlot``-compatible with:

```{r}
superb:::is.superbPlot.function("superbPlot.simple")
```

If ``TRUE``, then we are ready to use this layout, here with the
demo dataset ``TMB1964r``:

```{r, message=FALSE, echo=TRUE, results='hide', fig.show='hide'}
superbPlot(TMB1964r,
     WSFactor = "T(7)",      
     BSFactor = "Condition",
     variables = c("T1","T2","T3","T4","T5","T6","T7"),
     plotStyle = "simple", Quiet = TRUE
)
```


## Optional arguments

The above ``simple`` layout does not accept optional arguments. To
integrate optional arguments, one method is to insert graphic
directives inside the layers, e.g., inside ``geom_point``. 

A convenient method is with ``do.call`` and ``modifyList``, for example

```{r, eval = FALSE, message=FALSE, echo=FALSE, error=FALSE, results='hide'}
    do.call( geom_point, modifyList(
       list( size= 3 ##etc., the default directives##
       ), myownParams
    ))
```

A full example it therefore

```{r}
superbPlot.simpleWithOptions <- function(
    summarydata, xvar, groupingfac, addfactors, Debug, rawdata,
    myownParams = list()  ## add the optional arguments to the function
) {
    plot <- ggplot(
        data = summarydata, 
        mapping = aes_string( x = xvar, y = "center")
    ) +
    do.call( geom_point, modifyList(
       list( color ="black" ),
        myownParams
    )) + 
    geom_errorbar( mapping = aes_string(ymin = "center + lowerwidth", ymax = "center + upperwidth")  )+ 
    facet_grid( addfactors )
        
    return(plot)
}
superb:::is.superbPlot.function("superbPlot.simpleWithOptions")
```

```{r, message=FALSE, echo=TRUE, results='hide', fig.show='show', fig.cap="Figure 2: A simple figure with optional arguments"}
superbPlot(TMB1964r,
    WSFactor = "T(7)",      
    BSFactor = "Condition",
    variables = c("T1","T2","T3","T4","T5","T6","T7"),
    plotStyle = "simpleWithOptions", Quiet = TRUE,
    myownParams = list(size=12, color="red") ## here goes the optional arguments
)
```




## Getting Debug information

It is sometimes usefull to extract variables out of the function
when debugging the code. A useful function is to use ``runDebug``.
This function (shipped with ``suberb``) will display text and
transfer any variables you want into the global environnement.

```{r, eval=FALSE, message=FALSE, echo=TRUE, results='hide'}
runDebug( Debug, "Text to show",
  c("variable1", "variable2", "etc"),
  list( var1InTheFct, var2InTheFct, varetcInTheFct)
)
```

For example, the following will get the dataframes:

```{r, message=FALSE, echo=TRUE, results='hide', fig.show='hide'}
superbPlot.empty <- function(
    summarydata, xvar, groupingfac, addfactors, Debug, rawdata 
) {
    runDebug( Debug, "Getting the dataframes",
        c("summary","raw"), list(summarydata, rawdata))

    plot <- ggplot() # an empty plot        
    return(plot)
}
superbPlot(TMB1964r,
     WSFactor = "T(7)",      
     BSFactor = "Condition",
     variables = c("T1","T2","T3","T4","T5","T6","T7"),
     plotStyle = "empty", Quiet = TRUE, Debug = TRUE ## turn on Debug
)
```

(I inhibited the output, but you will get pages of debug information)
Browsing the information displayed, you will see at the end ``Getting the
dataframes`` followed by ``summary`` and ``raw``. These two variables
are not in the global environment and you can manipulate them. 
You can also use them in plotting functions, for example

```{r, message=FALSE, echo=TRUE, results='hide', fig.show='hide'}
superbPlot.simple(summary, "T", "Condition", ".~.", FALSE, raw)
``` 



## An example

In what follow, we create a toy example where the raw data
will be shown with smileys. Note that this example may not
work in Rstudio (see "limitation" on emojifont 
[page](https://cran.r-project.org/web/packages/emojifont/vignettes/emojifont.html)
)

We first need the ``emojifont`` library

```{r}
# install.packages("emojifont")
library(emojifont)
```

Then we define a ``"smiley"`` layout where the emojis are shown
with ``geom_text`` layer:

```{r}
superbPlot.smiley <- function(
    summarydata, xvar, groupingfac, addfactors, Debug, rawdata 
) {
    # the early part bears on summary data with variable "center"
    plot <- ggplot(
        data = summarydata, 
        mapping = aes_string(
            x = xvar, y = "center", 
            fill = groupingfac, 
            shape = groupingfac, 
            colour = groupingfac)
    ) +
    geom_point(position = position_dodge(width = .95)) +
    geom_errorbar( width = .6, position = position_dodge(.95), 
              mapping = aes_string(ymin = "center + lowerwidth", ymax = "center + upperwidth") 
    )+ 
    # this part bears on the rawdata only with variable "DV"
    geom_text(data = rawdata, 
              position = position_jitter(0.5),
              family="EmojiOne", label=emoji("smile"), size=3, 
              mapping=aes_string(x=xvar, y="DV", group = groupingfac)
    ) +
    facet_grid( addfactors )
        
    return(plot)
}
```

We check that it is a ``superbPlot``-compatible function:

```{r}
superb:::is.superbPlot.function("superbPlot.smiley")
```

It is all we need! It can be inserted in a ``superbPlot`` call:

```{r}
superbPlot(TMB1964r,
     WSFactor = "T(7)",      
     BSFactor = "Condition", 
     variables = c("T1","T2","T3","T4","T5","T6","T7"),
     plotStyle = "smiley", Quiet = TRUE
)
```


# References

