---
title: "Devising custom plot layouts"
output: 
  html_document:
    toc: TRUE
description: >
  superbPlot can generate either line plots, barcharts, or point plots. 
  However, additional layout can be programmed. In this vignette, we 
  show how to add custom layouts in superbPlot. As an example,
  we show how to implement additional layouts.
vignette: >
  %\VignetteIndexEntry{Devising custom plot layouts}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


```{r, echo = FALSE, message = FALSE, results = 'hide'}
cat("this will be hidden; use for general initializations.\n")
library(superb)
library(ggplot2)
```

``superbPlot`` comes with six built-in layouts for plotting your data.
However, it is possible to add additional, custom-made layouts. In this
vignette, we present rapidly the existing layouts, then show how
to supplement ``superb`` with your own layouts.


## The built-in plot layouts


When calling ``superbPlot``, you use the ``plotStyle = "layout"`` option
to indicate which layout you wish to use. Internally, ``superbPlot``
is calling a function whose name is ``superbPlot."layout"``. For example,
with ``plotStyle = "line"``, the plot is actually performed by the 
function ``superbPlot.line``.


The six layout available in  ``superbPlot`` package are :

* ``superbPlot.line`` : shows the results as points and lines, 

* ``superbPlot.point``: shows the results as points only, 

* ``superbPlot.bar`` : shows the results using bars, 

* ``superbPlot.pointjitter``: shows the results with points, and the raw data with jittered points, 

* ``superbPlot.pointjitterviolin``: also shows violin plot behind the jitter points, and 

* ``superbPlot.pointindividualline``: show the results with fat points, and individual results with thin lines, 

To determine if a certain function is ``superbPlot``-compatible, 
use the following function:

```{r}
superb:::is.superbPlot.function("superbPlot.line")
```

where you put between quote the name of a function. When devising your
own, custom-made function, it is a gook thing to check that it is
superbPlot-compatible.


## Illustrating the built-in layouts

To get a sense of the currently available layouts, we first generate 
a dataset composed of randomly generated scores mimicking a 3 $\times$
2 design with three degrees of Difficulties (as a between-group 
factor) and two days of testing (as a within-subject factor). It
is believed (and simulated) that all two factors have main effets on
the scores.

```{r}
testdata <- GRD(
    RenameDV   = "score", 
    SubjectsPerGroup = 25, 
    BSFactors  = "Difficulty(3)", 
    WSFactors  = "Day(day1, day2)",
    Population = list(mean = 65,stddev = 12,rho = 0.5),
    Effects    = list("Day" = slope(-5), "Difficulty" = slope(3) )
)
head(testdata)
```

For simplicity, we define a function whose arguments are the dataset and the layout:

```{r}
    mp <- function(data, style, ...) {
        superbPlot(data,
            WSFactor = "Day(2)",
            BSFactor = "Difficulty",
            variables = c("score.day1", "score.day2"),
           adjustments = list(purpose="difference", decorrelation="CM"),
            plotStyle = style,
            ...
        )+labs(title = paste("Layout is ''",style,"''",sep=""))
    }
```

Lets compute the plots will all six built-in layouts and show them

```{r, fig.width= 7, fig.height = 7, fig.cap = "Figure 1: Look of the six built-in layouts on the same random dataset"}
p1 <- mp(testdata, "bar")
p2 <- mp(testdata, "point")
p3 <- mp(testdata, "line")
p4 <- mp(testdata, "pointjitter" )
p5 <- mp(testdata, "pointjitterviolin")
p6 <- mp(testdata, "pointindividualline")

library(gridExtra)
grid.arrange(p1,p2,p3,p4,p5,p6,ncol=2)
```

For more controls, you can manually set the colors, the fills and/or
the shapes, as done here in a list:

```{r, fig.width= 7, fig.height = 7, fig.cap = "Figure 2: The six built-in template with ornamental styling added."}
ornate = list( 
    scale_colour_manual( name = "Difference", 
        labels = c("Easy", "Hard", "Unthinkable"), 
        values = c("blue", "black", "purple")) ,
    scale_fill_manual( name = "Difference", 
        labels = c("Easy", "Hard", "Unthinkable"), 
        values = c("blue", "black", "purple")) ,
    scale_shape_manual( name = "Difference", 
        labels = c("Easy", "Hard", "Unthinkable") ,
        values = c(0, 10, 13)) ,
    theme_bw(base_size = 9) ,
    labs(x = "Days of test", y = "Score in points" ),
    scale_x_discrete(labels=c("1" = "Former day", "2" = "Latter day"))
)
grid.arrange(
    p1+ornate, p2+ornate, p3+ornate,
    p4+ornate, p5+ornate, p6+ornate,
    ncol=2)
```

These are just a few examples. However, if these layouts do not fit
yours needs, it is possible to devise a custom-made layout and 
inform ``superbPlot`` to use it. To that end, see the instructions
below.


## Devising a custom-made plot layout

In a nutshell, the purpose of ``superbPlot`` is to 

* compile the summary information
(location of the summary statistic, upper width and lower width of 
the interval) and that, for each level of the factors; 

* applies all the adjustments needed in producing the summary;

* ``superbPlot`` also transform to discrete factors (using 
``as.factor``) all the factors. If some of the factors are
within-subject, the information is to be found in distinct column (
remember that ``superbPlot`` assumes a wide-format data frame); for
between-group factors, a column indicate the current group; 

* and finally, calls the plot function accepting pre-defined 
arguments

In devising your own plot function, it is important that (i) the 
function name begins with ``superbPlot.``; (ii) the
function accept very specific arguments with very precise names.

Here is the header for a function corresponding to a
plot style called, say, *foo* (``plotStyle = "foo"``):

```{r}
superbPlot.foo <- function(
    summarydata,
    xfactor,       
    groupingfactor,
    addfactors, 
    rawdata        
    # any optional argument you wish
) {
    plot <- ggplot() ## ggplot instructions...        
    return(plot)
}
```

In what follow, it is assumed that one factor is placed on the
horizontal axis (``xfactor``), another one is used to group the 
point (``groupingfactor``), and up to two additional
 factors will results in columns and rows of panels
 (``addfactors). (Of course, in devising your own template,
you may use different placement.) 


The arguments are:

* ``summarydata``: this data frame will contain the column ``center``
indicating the statistic's value, ``lowerwidth`` and ``upperwidth``
indicating how many units below and above ``center`` the error bar
extends. The data frame will also have columns for all the factors, 
and there will be as many lines as there are combinations of factors.

* ``xfactor`` is the factor to put on the horizontal axis;

* ``groupingfactor`` is the factor used to create groups of points;

* ``addfactors`` are up to two additional factors to create the 
rows and columns of panels. ``addfactors`` is formatted for facetting
(e.g., for factors "A" and "B", ``addfactors`` would be "A~B");

* ``rawdata`` : this data.frame contains the raw data with factors
being transformed ``as.factor`` and the dependent column being renamed
``DV``. When the data are in wide format, ``rawdata`` is reshaped to
long format.

* ``{optional arguments}`` can be used. They must be named here; when 
calling ``superbPlot``, any argument whose name match your optional
argument will be transmitted to your custom-made function.


## The simplest example


What follow is a simple example that will design a template 
that we will call ``simple``. This layout will display the
descriptive statistics and error bars. Everything will be
black and white (no color instruction) and superimposed
(no grouping instruction).

The result will be:

```{r, message=FALSE, echo=FALSE, fig.cap="Figure 3: Mean score with 95% confidence interval using the ``simple`` plot layout."}
superbPlot.simple <- function( summarydata, xfactor, groupingfactor, addfactors, rawdata ) {
    plot <- ggplot(
        data = summarydata,
        mapping = aes_string( x = xfactor, y = "center")
    ) +
    geom_point( ) +
    geom_errorbar( mapping = aes_string(ymin = "center + lowerwidth", ymax = "center + upperwidth")  )+
    facet_grid( addfactors )
       
    return(plot)
}
superbPlot(TMB1964r,
    WSFactor = "T(7)",
    BSFactor = "Condition",
    variables = c("T1","T2","T3","T4","T5","T6","T7"),
    plotStyle = "bar"
)
```

To make this plot, we design a function ``superbPlot.simple`` as:

```{r}
superbPlot.simple <- function(
    summarydata, xfactor, groupingfactor, addfactors, rawdata 
) {
    plot <- ggplot(
        data = summarydata, 
        mapping = aes_string( x = xfactor, y = "center")
    ) +
    geom_point( ) +
    geom_errorbar( mapping = aes_string(ymin = "center + lowerwidth", ymax = "center + upperwidth")  )+ 
    facet_grid( addfactors )
        
    return(plot)
}
```

The first instruction, ``ggplot`` defines the source data to be 
``summarydata`` with horizontal axis being in the string ``xfactor``
(this is the reason that ``mapping`` must be given with ``aes_string``).
The position of the descriptive statistics will be in ``"center"``.

The second instruction put points for each ``"center"``, and 
the third instruction places error bars. In that case, the ``ymin`` and 
``ymax`` information are contained in ``center+lowerwidth`` and
``center+upperwidth`` where ``lowerwidth`` and ``upperwidth`` are
in the summarydata dataframe.

The last instructions generates distinct panels for each level of the
remaining factors.
 
You can check that this function is ``superbPlot``-compatible with:

```{r}
superb:::is.superbPlot.function("superbPlot.simple")
```

If ``TRUE``, then we are ready to use this layout, here with the
demo dataset ``TMB1964r`` whose result was shown above in Figure 3:

```{r, message=FALSE, echo=TRUE, results='hide', fig.show='hide'}
superbPlot(TMB1964r,
     WSFactor = "T(7)",      
     BSFactor = "Condition",
     variables = c("T1","T2","T3","T4","T5","T6","T7"),
     plotStyle = "simple"
)
```


## Optional arguments

The above ``simple`` layout does not accept optional arguments. To
integrate optional arguments, one method is to insert graphic
directives inside the layers, e.g., inside ``geom_point``. 

A convenient method is with ``do.call`` and ``modifyList``, for example:

```{r, eval = FALSE, message=FALSE, echo=TRUE, error=FALSE, results='hide'}
    do.call( geom_point, modifyList(
       list( size= 3 ##etc., the default directives##
       ), myownParams
    ))
```

A full example it therefore

```{r}
superbPlot.simpleWithOptions <- function(
    summarydata, xfactor, groupingfactor, addfactors, rawdata,
    myownParams = list()  ## add the optional arguments to the function
) {
    plot <- ggplot(
        data = summarydata, 
        mapping = aes_string( x = xfactor, y = "center")
    ) +
    do.call( geom_point, modifyList(
       list( color ="black" ),
        myownParams
    )) + 
    geom_errorbar( mapping = aes_string(ymin = "center + lowerwidth", ymax = "center + upperwidth")  )+ 
    facet_grid( addfactors )
        
    return(plot)
}
superb:::is.superbPlot.function("superbPlot.simpleWithOptions")
```

```{r, message=FALSE, echo=TRUE, results='hide', fig.show='show', fig.cap="Figure 4: A simple figure with optional arguments"}
superbPlot(TMB1964r,
    WSFactor = "T(7)",      
    BSFactor = "Condition",
    variables = c("T1","T2","T3","T4","T5","T6","T7"),
    plotStyle = "simpleWithOptions", 
    myownParams = list(size=12, color="red") ## here goes the optional arguments
)
```




## Getting Debug information

It is sometimes useful to extract variables out of the function
when debugging the code. A useful function is to use ``runDebug``.
This function (shipped with ``suberb``) will display text and
transfer any variables you want into the global environment.

```{r, eval=FALSE, message=FALSE, echo=TRUE, results='hide'}
options(superb.debug = 'all')
runDebug( 'where are we?', "Text to show when we get there",
  c("variable1", "variable2", "etc"),
  list( "var1InTheFct", "var2InTheFct", "varetcInTheFct")
)
```

For example, the following will get the dataframes:

```{r, message=FALSE, echo=TRUE, results='hide', fig.show='hide'}
superbPlot.empty <- function(
    summarydata, xfactor, groupingfactor, addfactors, rawdata 
) {
    runDebug( 'inempty', "Dumping the two dataframes",
        c("summary","raw"), list(summarydata, rawdata))

    plot <- ggplot() # an empty plot        
    return(plot)
}
options(superb.debug = 'inempty')  ## turn on Debug when reaching 'inempty'
superbPlot(TMB1964r,
     WSFactor = "T(7)",      
     BSFactor = "Condition",
     variables = c("T1","T2","T3","T4","T5","T6","T7"),
     plotStyle = "empty" 
)
```

You will see  ``Dumping the two dataframes``
 followed by ``summary`` and ``raw``. These two variables
are now in the global environment and you can manipulate them. 
You can also use them in testing your plotting functions, for example

```{r, message=FALSE, echo=TRUE, results='hide', fig.show='hide'}
superbPlot.simple(summary, "T", "Condition", ".~.", raw)
``` 



## An example

In what follow, we create a toy example where the raw data
will be shown with smileys. Note that this example may not
work in Rstudio (see "limitation" on emojifont 
[page](https://CRAN.R-project.org/package=emojifont)
)

We first need the ``emojifont`` library

```{r}
# install.packages("emojifont")
library(emojifont)
```

Then we define a ``"smiley"`` layout where the emojis are shown
with ``geom_text`` layer:

```{r}
superbPlot.smiley <- function(
    summarydata, xfactor, groupingfactor, addfactors, rawdata 
) {
    # the early part bears on summary data with variable "center"
    plot <- ggplot(
        data = summarydata, 
        mapping = aes_string(
            x = xfactor, y = "center", 
            fill = groupingfactor, 
            shape = groupingfactor, 
            colour = groupingfactor)
    ) +
    geom_point(position = position_dodge(width = .95)) +
    geom_errorbar( width = .6, position = position_dodge(.95), 
              mapping = aes_string(ymin = "center + lowerwidth", ymax = "center + upperwidth") 
    )+ 
    # this part bears on the rawdata only with variable "DV"
    geom_text(data = rawdata, 
              position = position_jitter(0.5),
              family="EmojiOne", label=emoji("smile"), size=6, 
              mapping=aes_string(x=xfactor, y="DV", group = groupingfactor)
    ) +
    facet_grid( addfactors )
        
    return(plot)
}
```

We check that it is a ``superbPlot``-compatible function:

```{r}
superb:::is.superbPlot.function("superbPlot.smiley")
```

It is all we need! It can be inserted in a ``superbPlot`` call:

```{r, fig.width= 7, fig.height = 6, fig.cap = "Figure 5: smile!" }
superbPlot(TMB1964r,
     WSFactor = "T(7)",      
     BSFactor = "Condition", 
     variables = c("T1","T2","T3","T4","T5","T6","T7"),
     plotStyle = "smiley"
)
```


